Sequential rank agreement methods for comparison of lists

### 2015-08-05 latex/paper.Rnw
Updating application section

We use the following four classification methods: Random Forest \citep{breiman2001random} implemented in the R package \texttt{randomForest} \citep{liaw2002classification}, logistic Lasso \citep{tibshirani1996regression} and Ridge regression \citep{segerstedt1992ordinary} both implemented in the R package \texttt{glmnet} \citep{Friedman:2010aa}, and Partial Least Squares Discriminant Analysis (PLS-DA) \citep{boulesteix2004pls} implemented in the R package \texttt{caret} \citep{Jed-Wing:2014aa}.

@@

All four models depend on tuning parameters. These are estimated by 20 times repeated 5-fold cross-validation on the training data in each iteration. The implementation of Lasso and Ridge regression in the \texttt{glmnet} package offers three different cross-validated optimization criteria for the penalty parameter: total deviance, classification accuracy and Area Under ROC. We apply all three criteria to our data to investigate their effect on the agreement.

\begin{figure}[tbp]
\begin{center}
\includegraphics[width=.5\textwidth]{pics/riskAgreementPlot}%
\includegraphics[width=.5\textwidth]{pics/predictorAgreementPlot}
\end{center}
 \caption{Sequential rank agreement of predicted risk probabilities for the DACOVA data (left) and estimated predictor importance (right) based on 1000 sub-samples of the MALOVA training data.}
 \label{fig:app1}
\end{figure}

Figure \ref{fig:app1} (left) shows the sequential rank agreement applied to the estimated risk probabilities of the $113$ patients in the DACOVA data set obtained from the $1000$ sub-samples of the MALOVA data, and Figure \ref{fig:app1} (right) shows the sequential rank agreement of the estimated importance of the 5000 predictors. For clarity we only show the agreement for the latter up to list depth 600 where all the curves have approximately become stable. We note here that the predictor importance estimated with the Lasso model is capable of producing censored lists depending on the value of the penalty parameter. With the deviance criterion the median number of non-zero coefficients was 33 (range 18 to 52). In order to calculate the sequential rank agreement we therefore applied Algorithm \ref{sra-algorithm} with $B = 1000$ and $P=5000$. The effect of the censoring caused by the sparsity of the Lasso is clearly visible in the agreement curves.

For most of the sequential rank agreement curves we see as expected that they start low followed by an increase until they approximately become constant.  This has the interpretation that the classification methods show higher agreement with respect to the highest ranked items (the most high-risk patients and the most important predictors respectively) then for the items deeper in the lists across the different sub-samples. The change-point where the curves become approximately constant is the list depth where the ranks of the remaining items become almost random. The only exceptions are the predictor importance agreements using the Ridge models and the risk probability agreement using the Lasso model with the AUC criterion which show higher disagreement in the top of the lists. The reason for this in the Lasso case is that optimizing with respect to the AUC criterion tends to favor a very high value of the penalty parameter causing only a single predictor to be selected in each of the 1000 iterations. This results in a lack of generalizability to the DACOVA data which gives rise to the higher disagreement in the predicted risk probabilities. In the extreme case where the penalty becomes so high that none of the predictors are selected by the Lasso, the sequential rank agreement for the predicted probabilities becomes undefined since all the ranks will be ties.

From Figure \ref{fig:app1} it can also be seen that some of the methods show better agreement with respect to the predicted probabilities than for ranking  the importance of the predictors and vice versa. Ridge regression performs very well for the risk predictions but is outperformed by PLS-DA for the predictor importances. On the other hand, PLS-DA appears to be similarly worse for the risk predictions compared to Ridge regression. The Random Forest method seems to perform equally well in both scenarios. Lasso regression (except with the AUC criterion as discussed previously) performs very well at ranking the high-risk patients but very poorly for the importance of the predictors. This reason for this is the high auto-correlation between the intensities in the mass spectra which leads to collinearity issues in the regression models. It is well-known that variable selection with the Lasso penalty does not preform very well when the predictors are highly correlated. This collinearity does, however, not affect the risk predictions, as seen on the left part of the figure, since it is not so important which specific variable that get selected from a group of highly correlated predictors when the purpose is risk predictions.

Finally, it would appear that Ridge regression with the AUC criterion for the penalty parameter achieves the best performance with respect to the stability of ranking the individual predicted risk probabilities. It must, however, be stressed that the sequential rank agreement is only concerned with whether similar orderings of the patients is obtained under perturbations of the training data. It is not concerned with whether such an ordering is the correct one. This is similar to the usual distinction between accuracy and precision. Looking at the distributions of the AUCs across the $1000$ iterations for each of the models with respect to the known class labels of the DACOVA data we see that the PLS-DA model attained the best AUC with a median value of $0.71$ (IQR = 0.07) while the Ridge regression with AUC criterion has a median AUC of $0.49$ (IQR = 0.02). This means that while Ridge regression optimized with respect to the AUC criterion achieves the best rank agreement it performs similar to a random coin toss with respect to classification on the DACOVA data. Both concerns are of course important in practice and for this data set it appears that PLS-DA is the overall best method taking both agreement and classification into account.

>>>>

We use the following four classification methods: Random Forest \citep{breiman2001random} implemented in the R package \texttt{randomForest} \citep{liaw2002classification}, logistic Lasso \citep{tibshirani1996regression} and Ridge regression \citep{segerstedt1992ordinary} both implemented in the R package \texttt{glmnet} \citep{friedman2010regularization}, and Partial Least Squares Discriminant Analysis (PLS-DA) \citep{boulesteix2004pls} implemented in the R package \texttt{caret} \citep{Jed-Wing:2014aa}.

@@

All four models depend on tuning parameters. We estimate these by a 20 times repeated 5-fold cross-validation on the training data in each iteration. The implementation of Lasso and Ridge regression in the \texttt{glmnet} package offers three different cross-validated optimization criteria for the penalty parameter: total deviance, classification accuracy and Area Under ROC. We apply all three criteria to our data to investigate their effect on the agreement.

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=.5\textwidth]{pics/riskAgreementPlot}%
\includegraphics[width=.5\textwidth]{pics/predictorAgreementPlot}
\end{center}
 \caption{Sequential rank agreement of predicted risk probabilities for the DACOVA data (left) and predictor importance (right) based on 1000 sub-samples of the MALOVA training data.}
 \label{fig:app1}
\end{figure}

Figure \ref{fig:app1} (left) shows the sequential rank agreement applied to the estimated risk probabilities of the $113$ patients in the DACOVA data set obtained from the $1000$ sub-samples of the MALOVA data, and Figure \ref{fig:app1} (right) shows the sequential rank agreement of the estimated importance of the 5000 predictors. For clarity we only show the agreement for the latter up to list depth 600 where all the curves have approximately become stable. We note that the predictor importances estimated with the Lasso model are capable of producing censored lists depending on the value of the penalty parameter. Under the deviance optimization criterion the median number of non-zero coefficients was 33 (range 18 to 52). In order to calculate the sequential rank agreement we therefore applied Algorithm \ref{sra-algorithm} with $B = 1000$ and $P=5000$. The effect of the censoring caused by the sparsity of the Lasso is clearly visible in the agreement curves.

For most of the sequential rank agreement curves we see, as expected, that they start low followed by an increase until they approximately become constant.  This has the interpretation that the classification methods show higher agreement with respect to the highest ranked items (the most high-risk patients and the most important predictors respectively) than for the items deeper in the lists across the different sub-samples. The change-points where the curves become approximately constant are the list depths where the ranks of the remaining items become close to random.

The only exceptions to the expected shape of the agreement curves are the predictor importance agreements from the Ridge models and the risk prediction agreement from the Lasso model with the AUC criterion. They all show higher disagreement in the top of the lists followed by a decrease. The explanations for their behavior is different for the two methods. For the Lasso model the reason is that optimizing the penalty parameter with respect to the AUC criterion tends to favor a very high penalty value causing only a single predictor to be selected in each of the 1000 iterations. This results in a lack of generalizability to the DACOVA data which gives rise to the higher disagreement in the predicted risk probabilities. In the extreme case where the penalty becomes so high that none of the predictors are selected by the Lasso, the sequential rank agreement for the predicted probabilities becomes undefined since all the ranks will be ties.

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=.5\textwidth]{pics/ridgeCensoring}%
\includegraphics[width=.5\textwidth]{pics/aucPlot}
\end{center}
 \caption{Sequential rank agreement for Ridge regression with an artificial censoring on the absolute value of the regression coefficients (left) and box plots of AUC values across the $1000$ sub-samples with respect to the known class labels of the DACOVA data (right).}
 \label{fig:app2}
\end{figure}

The reason behind the behavior of the sequential rank agreement of predictor importances estimated by the Ridge models is more subtle. Looking at the distribution of the regression coefficients it is seen that a large proportion of them are numerically very close to zero and of almost equal absolute value. This implies that when they are ranked according to their magnitude, their actual order becomes more uncertain and more close to a random permutation. This problem can be alleviate by introducing an artificial censoring depending on the absolute value of the coefficients. Figure \ref{fig:app2} (left) shows the sequential rank agreement curve of the rankings of the predictor importances from the Ridge model using the deviance criterion (same curve as in figure \ref{fig:app1} (left)) as well as an agreement curve where the absolute value of the regression coefficients have been censored at the $0.1\%$ quantile. In our application the median number of censored predictors across the 1000 sub-samples is 5 (range 1 to 11). Even though the number of censored predictors is very small compared to the total number of predictors, the effect on the sequential rank agreement is substantial as it has now flipped to show higher agreement in the top of the lists as was expected.

Returning to Figure \ref{fig:app1}, it can further be seen that some of the methods show better agreement with respect to the predicted probabilities than for ranking the importance of the predictors and vice versa. Ridge regression performs very well for the risk predictions, but PLS-DA performs better for the predictor importances and similarly worse for the risk predictions. The Random Forest method seems to perform equally well in both scenarios. Lasso regression (except with the AUC criterion) performs very well at ranking the high-risk patients but very poorly at ranking the importance of the predictors. This reason for this is the high auto-correlation between the intensities in the mass spectra which leads to collinearity issues in the regression models. It is well-known that variable selection with the Lasso penalty does not preform very well when the predictors are highly correlated. This collinearity does, however, not affect the agreement of the risk predictions, as seen on the left part of the figure, since it is not so important which specific variable that gets selected from a group of highly correlated predictors when the purpose is risk predictions.

Finally, it would appear that Ridge regression with the AUC criterion for the penalty parameter achieves the best performance with respect to the stability of ranking the individual predicted risk probabilities. It must, however, be stressed that the sequential rank agreement in this application is only concerned with whether similar orderings of the patients is obtained under perturbations of the training data. It is not concerned with whether such an ordering is the correct one. This is similar to the usual distinction between accuracy and precision. Looking at the distributions of the AUCs illustrated in Figure \ref{fig:app2} (left) across the $1000$ sub-samples for a selection of the models with respect to the known class labels of the DACOVA data, we see that PLS-DA attains the highest AUC with a median value of $0.71$ while the Ridge model with the AUC criterion attains a median AUC of $0.49$. This implies that while Ridge regression optimized with respect to the AUC criterion achieves the best sequential rank agreement it performs similar to a random coin toss with respect to classifying the DACOVA patients. Both concerns are of course important in practice and for this data set it appears that PLS-DA is the overall best method taking both agreement and classification accuracy into account.
